# GEV responses

In this tutorial, we illustrate how to set up a distributional regression model with the generalized extreme value distribution as a response distribution. First, we simulate some data in R:

- The location parameter ($\mu$) is a function of an intercept and a non-linear covariate effect.
- The scale parameter ($\sigma$) is a function of an intercept and a linear effect and uses a log-link.
- The shape or concentration parameter ($\xi$) is a function of an intercept and a linear effect.

After simulating the data, we can configure the model with a single call to the `rliesel::liesel()` function.

```{r}
library(rliesel)
library(VGAM)

use_liesel_venv()
set.seed(1337)

n <- 1000

x0 <- runif(n)
x1 <- runif(n)
x2 <- runif(n)

y <- rgev(
  n,
  location = 0 + sin(2 * pi * x0),
  scale = exp(-3 + x1),
  shape = 0.1 + x2
)

plot(y)

model <- liesel(
  response = y,
  distribution = "GeneralizedExtremeValue",
  predictors = list(
    loc = predictor(~ s(x0)),
    scale = predictor(~ x1, inverse_link = "Exp"),
    concentration = predictor(~ x2)
  )
)
```

Now, we can continue in Python and use the `lsl.dist_reg_mcmc()` function to set up a sampling algorithm with IWLS kernels for the regression coefficients ($\boldsymbol{\beta}$) and a Gibbs kernel for the smoothing parameter ($\tau^2$) of the spline. Note that we need to set $\beta_0$ for $\xi$ to 0.1 manually, because $\xi = 0$ breaks the sampler.

```{python}
import liesel.liesel as lsl
import jax.numpy as jnp

model = r.model

# concentration = 0 seems to break the sampler
model.nodes["concentration_p0_beta"].value = jnp.array([0.1, 0.0])

builder = lsl.dist_reg_mcmc(model, seed=42, num_chains=4)
builder.set_duration(warmup_duration=1000, posterior_duration=1000)

engine = builder.build()
engine.sample_all_epochs()
```

Some tabular summary statistics of the posterior samples:

```{python results="asis"}
import liesel.goose as gs

results = engine.get_results()
summary = gs.summary(results)

summary.loc["loc_p0_beta"]
summary.loc["loc_np0_tau2"]

summary.loc["scale_p0_beta"]

summary.loc["concentration_p0_beta"]
```

And the corresponding trace plots:

```{python}
fig = gs.plot_trace(results, "loc_p0_beta")
fig = gs.plot_trace(results, "loc_np0_tau2")
fig = gs.plot_trace(results, "loc_np0_beta")

fig = gs.plot_trace(results, "scale_p0_beta")

fig = gs.plot_trace(results, "concentration_p0_beta")
```

We need to reset the index of the `summary` data frame before we can transfer it to R.

```{python}
summary = summary.reset_index()
```

After transferring the `summary` data frame to R, we can process it with packages like dplyr and ggplot2. Here is a visualization of the estimated spline vs. the true function:

```{r}
library(dplyr)
library(ggplot2)
library(reticulate)

summary <- py$summary

beta <- summary %>%
  filter(index == "loc_np0_beta") %>%
  group_by(param_index) %>%
  summarize(mean = mean(mean)) %>%
  ungroup()

beta <- beta$mean
X <- py_to_r(model$nodes$loc_np0_X$value)
estimate <- X %*% beta

true <- sin(2 * pi * x0)

ggplot(data.frame(x0 = x0, estimate = estimate, true = true)) +
  geom_line(aes(x0, estimate), color = palette()[2]) +
  geom_line(aes(x0, true), color = palette()[4]) +
  ggtitle("Estimated spline (red) vs. true function (blue)") +
  ylab("f") +
  theme_minimal()
```
